---
title: "Theropod Teeth- Higher taxonomic level"
author: "Carolina Marques"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    fig_caption: yes
    force_captions: yes
    highlight: pygments
    number_sections: yes
    theme: cerulean
    df_print: paged
    toc: yes
    toc_depth: 4
    toc_float: yes
bibliography: biblio.bib  
link-citations: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r,include=FALSE, warning=FALSE,message=FALSE}
library("ggplot2")
library(corrplot)
library("devtools")
library(readxl)
library(data.table)
library("RColorBrewer") 
library(kableExtra)
library(dplyr)
library(mgcv)
library(mgcViz)
```



# Reading the data 


```{r, warning=FALSE,message=FALSE}

data <- read_xlsx("Crown measurement dataset Kem Kem theropods.xlsx")

data[data == "?"] <- NA
data[data == "~"] <- NA
data[data == "/"] <- NA
# Remove "?" from all columns
data[] <- lapply(data, function(x) {
  gsub("\\? ","", x)
})

data[] <- lapply(data, function(x) {
  gsub("absent",0, x)
})

# Remove "?" from all columns
data[] <- lapply(data, function(x) {
  gsub("\\?","", x)
})


# Remove "?" from all columns
data[] <- lapply(data, function(x) {
  gsub("\\>","", x)
})

# Remove "?" from all columns
data[] <- lapply(data, function(x) {
  gsub("\\<","", x)
})
# Remove "?" from all columns
data[] <- lapply(data, function(x) {
  gsub(">","", x)
})

# Remove "?" from all columns
data[] <- lapply(data, function(x) {
  gsub("<","", x)
})
# Remove "?" from all columns
data[] <- lapply(data, function(x) {
  gsub("\\~","", x)
})

# Remove "?" from all columns
data[] <- lapply(data, function(x) {
  gsub("~","", x)
})

# Remove "?" from all columns
data[] <- lapply(data, function(x) {
  gsub(":","", x)
})
# Remove "?" from all columns
data[] <- lapply(data, function(x) {
  gsub(";","", x)
})
data$LIF<- ifelse(data$LIF=="6-7",6.5,
                  ifelse(data$LIF=="5-6",5.5,
                         ifelse(data$LIF=="4-5",4.5,
                                ifelse(data$LIF=="3-4","3.5",
                                                           ifelse(data$LIF=="11 or 12",11.5,
                                                                  ifelse(data$LIF=="10-13",12,data$LIF))))))
data$CH<-data$CH...22 
data<- data %>% select(-CH...60,-CH...22,-`(DDL/CH)*100`)
data1<-data
data<-data[,-c(1,2,3,4,6:14,16:19)]#until 19

data$`TransvUndu`<-ifelse(data$`Transv. Undu.`!=0 & !is.na(data$`Transv. Undu.`),1,data$`Transv. Undu.`)

data$`Interdentsulci`<-ifelse(data$`Interdent. sulci`!=0 & !is.na(data$`Interdent. sulci`),1,data$`Interdent. sulci`)

data$LAF<-ifelse(data$LAF=="6-7",6.5,data$LAF)

data$CTU1 <- sub(".*?(\\d+).*", "\\1", data$CTU)

data<- data %>% select(-CTU,-`Interdent. sulci`,-`Transv. Undu.`)

# Convert columns to numeric, then create log-transformed columns
data <- data %>%
  mutate(across(3:ncol(data), as.numeric)) %>%
  mutate(across(3:ncol(data), log, .names = "Log_{.col}"))

data$CladeToothtype<-as.factor(data$CladeToothtype)
#data$Epoch<-ifelse(data$Epoch=="'Middle Cretaceous'","Middle Cretaceous", data$Epoch)
#data$Epoch<-as.factor(data$Epoch)


data<-data.frame(data)


```


# Checking the data

## Summary of the table


```{r}
summary(data)
```


## Check first rows

```{r}
head(data)
```

## Tables

### Genus / Taxa Table

```{r}
taxa<-table(data$CladeToothtype)
data_taxa<-data.frame(taxa)
data_taxa<-data_taxa[order(data_taxa$Freq, decreasing = T), ]
data_taxa$ID<-1:nrow(data_taxa)
data_taxa
```


# Data Processing


## Removing columns that have more more missing value


```{r}
# Count the number of missing values in each column
missing_counts <- colSums(is.na(data))

# Remove columns with more than 15% missing values
data1_cleaned <- data[, missing_counts <= nrow(data)*0.15]

```


## Removing rows with NA values

```{r}
# Remove rows with any NA values
data1_cleaned <- na.omit(data1_cleaned)
data1_cleaned

```


## Subtracting the observations for the Taxa that have least observations



```{r}
lennn<-(ncol(data1_cleaned)-2)/2
taxa1<-table(data1_cleaned$CladeToothtype)
data1_cleanedd<-data.frame(taxa1)
data1_cleanedd<-data1_cleanedd[order(data1_cleanedd$Freq, decreasing = T), ]
data1_cleanedd$ID<-1:nrow(data1_cleanedd)
data1_cleanedd$CladeToothtype<-data1_cleanedd$Var1

data1_cleaned1<-data1_cleanedd[data1_cleanedd$Freq>lennn,]

data1_cleaned<-data1_cleaned[data1_cleaned$CladeToothtype%in%unique(data1_cleaned1$CladeToothtype),]

summary(data1_cleaned)
```


## Spliting the Log variables and the original

```{r}
# Select variables that contain "log" and the first column
selected_cols <- c(1,2, grep("Log", names(data1_cleaned)))

# Subset the data frame
data_log <- data1_cleaned[, selected_cols]

data_log

names(data_log)[-1]  <- gsub(" ", "_", names(data_log)[-1] )


# Identify columns that contain "log"
log_cols <- grep("Log", names(data1_cleaned))

# Include the first column
cols_to_keep <- setdiff(1:ncol(data1_cleaned), log_cols)

# Ensure the first column is included
cols_to_keep <- union(1, cols_to_keep)

# Subset the data frame
data_original <- data1_cleaned[, cols_to_keep]

data_original
```

# Visualization



## Original data

```{r}
correlation_matrix <- cor(data_original[, -c(1, 2)])
# Plot correlation matrix
corrplot(correlation_matrix, method = "color", type = "lower", 
         addCoef.col = "black", 
         tl.col = "black", 
         tl.srt = 45, 
         diag = FALSE, 
         order = "hclust", 
         col = colorRampPalette(c("blue", "white", "red"))(200))
```


## Log data


```{r}
correlation_matrix <- cor(data_log[, -c(1, 2)])
# Plot correlation matrix
corrplot(correlation_matrix, method = "color", type = "lower", 
         addCoef.col = "black", 
         tl.col = "black", 
         tl.srt = 45, 
         diag = FALSE, 
         order = "hclust", 
         col = colorRampPalette(c("blue", "white", "red"))(200))
```

```{r}
colnames(data_original)

# Filter out taxa with less than 10 observations
data_log <- data_log %>%
  group_by(CladeToothtype) %>%                   # Group by 'Taxa' column
  filter(n() >= 10) %>%                # Keep only groups with 10 or more observations
  ungroup()                            # Ungroup after filtering
# Filter out taxa with less than 10 observations
data_original <- data_original %>%
  group_by(CladeToothtype) %>%                   # Group by 'Taxa' column
  filter(n() >= 10) %>%                # Keep only groups with 10 or more observations
  ungroup() 
# Get the count of each unique value in the column
category_counts <- table(data_log$CladeToothtype)

# Filter unique values that have more than 0 observations
unique_values <- names(category_counts[category_counts > 0])

# Print the result
#print(unique_values)
```

# Merging data clade with tooth to add more classes

```{r}
data1$CladeToothtype<-ifelse(!(data1$CladeToothtype%in%unique_values),data1$Clade,data1$CladeToothtype)

data1<-data1[,-c(1,2,3,4,6:14,16:19)]#until 19
#clade: 4, taxa:2, teethtaxa: 3, cladetteth: 5, epoch:15
#data<-inner_join(dd,data)
#data<-data[!duplicated(data),]

data1$`TransvUndu`<-ifelse(data1$`Transv. Undu.`!=0 & !is.na(data1$`Transv. Undu.`),1,data1$`Transv. Undu.`)

data1$`Interdentsulci`<-ifelse(data1$`Interdent. sulci`!=0 & !is.na(data1$`Interdent. sulci`),1,data1$`Interdent. sulci`)

data1$LAF<-ifelse(data1$LAF=="6-7",6.5,data1$LAF)

data1$CTU1 <- sub(".*?(\\d+).*", "\\1", data1$CTU)

data1<- data1 %>% select(-CTU,-`Interdent. sulci`,-`Transv. Undu.`)

# Convert columns to numeric, then create log-transformed columns
data1 <- data1 %>%
  mutate(across(3:ncol(data1), as.numeric)) %>%
  mutate(across(3:ncol(data1), log, .names = "Log_{.col}"))

data1$CladeToothtype<-as.factor(data1$CladeToothtype)
#data1$Epoch<-ifelse(data1$Epoch=="'Middle Cretaceous'","Middle Cretaceous", data1$Epoch)
#data1$Epoch<-as.factor(data1$Epoch)


#data$Taxa<-as.factor(paste0(data$`Taxa (Genus)`,data$Maturity,sep=" "))
data1<-data.frame(data1)



# Count the number of missing values in each column
missing_counts <- colSums(is.na(data1))

# Remove columns with more than 15% missing values
data1_cleaned <- data1[, missing_counts <= nrow(data1)*0.15]

# Remove rows with any NA values
data1_cleaned <- na.omit(data1_cleaned)
data1_cleaned

lennn<-(ncol(data1_cleaned)-2)/2
taxa1<-table(data1_cleaned$CladeToothtype)
data1_cleanedd<-data.frame(taxa1)
data1_cleanedd<-data1_cleanedd[order(data1_cleanedd$Freq, decreasing = T), ]
data1_cleanedd$ID<-1:nrow(data1_cleanedd)
data1_cleanedd$CladeToothtype<-data1_cleanedd$Var1

data1_cleaned1<-data1_cleanedd[data1_cleanedd$Freq>lennn,]

data1_cleaned<-data1_cleaned[data1_cleaned$CladeToothtype%in%unique(data1_cleaned1$CladeToothtype),]

summary(data1_cleaned)

# Select variables that contain "log" and the first column
selected_cols <- c(1,2, grep("Log", names(data1_cleaned)))

# Subset the data frame
data_log <- data1_cleaned[, selected_cols]

data_log

names(data_log)[-1]  <- gsub(" ", "_", names(data_log)[-1] )


# Identify columns that contain "log"
log_cols <- grep("Log", names(data1_cleaned))

# Include the first column
cols_to_keep <- setdiff(1:ncol(data1_cleaned), log_cols)

# Ensure the first column is included
cols_to_keep <- union(1, cols_to_keep)

# Subset the data frame
data_original <- data1_cleaned[, cols_to_keep]

data_original

colnames(data_original)

# Filter out taxa with less than 10 observations
data_log <- data_log %>%
  group_by(CladeToothtype) %>%                   # Group by 'Taxa' column
  filter(n() >= 10) %>%                # Keep only groups with 10 or more observations
  ungroup()                            # Ungroup after filtering
# Filter out taxa with less than 10 observations
data_original <- data_original %>%
  group_by(CladeToothtype) %>%                   # Group by 'Taxa' column
  filter(n() >= 10) %>%                # Keep only groups with 10 or more observations
  ungroup() 
# Get the count of each unique value in the column
category_counts <- table(data_log$CladeToothtype)

# Filter unique values that have more than 0 observations
unique_values1 <- names(category_counts[category_counts > 0])

```

# Saving the clean dataset

```{r}
write.csv(data_log,"teeth_data_log_clade_epoch1.csv", row.names = FALSE)
write.csv(data_original,"teeth_data_clade_epoch1.csv", row.names = FALSE)
```

# Summary tables

```{r}
library(readxl)
data <- read_xlsx("Crown measurement dataset Kem Kem theropods.xlsx")

table(data$Country)

table(data$Maturity)

table(data$Epoch)

```


